---
title: "Macrocystis WGS Project"
author: "Meaghan Clark"
date: "2026-02-24"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Summary 

Project begun Februray 2026. Goal is to combine multiple *Macrocystis* whole genome resequencing data sets to learn about structure, inbreeding, and local adaptation. 


## Directory structure

```
MpyrAdapt/
├── data/
│   ├── raw/          # symbolic links to raw sequencing data
│   ├── processed/    # trimmed, filtered, aligned data
│   └── reference/    # symbolic link to reference genome
├── scripts/
│   ├── 0. sra_download.sbatch
│   ├── 1. make_smbolic_links.sbatch
│   ├── 1.5 make_smbolic_links_CCGP.sbatch
│   ├── 2. trim_reads.sbatch
│   ├── 3. make_alignment_key.sbatch
│   ├── 4. wrapper-align_to_genome.sh
│   ├──── align_to_genome.sbatch
│   ├── 5. wrapper-run_qualimap.sh
│   ├──── run_qualimap.sbatch
│   ├── 6. summarize_qualimap.sbatch
│   ├── 7. merge_bams.sbatch
│   ├── 7. wrapper-run_bamstats.sh
│   ├──── run_bamstats.sbatch
│   ├── 8. make_bamQC_histograms.sbatch
│   ├──── make_bamQC_histograms.R
│   ├── 9. wrapper-generate_PF_masks.sh
│   ├──── generate_PF_masks.sbatch
│   ├── 10. wrapper-call_snps.sh
│   ├──── call_snps.sbatch
│   ├── 11. wrapper-normalize_snps.sh
│   ├──── normalize_snps.sbatch
│   ├── 12. wrapper-annotate_VCF.sh
│   ├──── annotate_VCF.sbatch
│   ├── 
│   ├── 
│   ├── 
│   ├── 

│   └── 
├── 
│   ├── 
│   └── 
├── logs
└── MpyrAdapt_notebook.md
```

## Metadata 

##### Site keys from Alberto data 

---
AM	Andrew Molera	36.283333	-121.876392
MO	Monterey	36.617145	-121.895693
PL	Point Loma	32.678617	-117.253367
PP	Santa Cruz	36.952702	-121.9711
PV	Palos Verdes	33.720889	-118.336278
SC	Santa Cruz Island	34.032411	-119.524502
SM	San Miguel Island	34.051883	-120.33885
---

#### Gonzalez et al. PRJNA938791  
[Whole-genome sequencing distinguishes the two most common giant kelp ecomorphs](https://academic.oup.com/evolut/article/77/6/1354/7079449)

---
Site name	Site abbreviation	Coordinates	Hemisphere	Morph	Collected by
Stillwater Cove	STL	36.565058, -121.943553	North	Pyrifera and Integrifolia	S. T. Gonzalez
Cayucos	CAY	35.446456, -120.932394	North	Pyrifera	S. T. Gonzalez
Arroyo Quemado	AQ	34.47, -120.119	North	Pyrifera	F. Alberto
Catalina Island	CI	33.462, -118.511	North	Pyrifera	F. Alberto
Camp Pendleton	CB	33.157, -117.36	North	Pyrifera	F. Alberto
Playa Blanca	PLB	-28.1851111, -71.1651667	South	Integrifolia	S. T. Gonzalez
Punta Choros	PCH	-29.245712, -71.466174	South	Integrifolia	S. T. Gonzalez
Totorolillo	TOT	-32.022839, -71.508658	South	Integrifolia	S. T. Gonzalez
Las Docas	DOC	-33.139851, -71.708218	South	Integrifolia	S. T. Gonzalez
Algarrobo	ALG	-33.360022, -71.668728	South	Integrifolia	S. T. Gonzalez
Las Monjas	MON	-33.492179, -71.642485	South	Integrifolia	S. T. Gonzalez
La Boca	BOC	-33.903175, -71.834772	South	Pyrifera	S. T. Gonzalez
Tumbes	TUM	-36.623731, -73.093019	South	Pyrifera	S. T. Gonzalez
Chome	CHO	-36.806467, -73.177088	South	Pyrifera	S. T. Gonzalez
Bahía Mansa	BAM	-40.581556, -73.738197	South	Pyrifera	S. T. Gonzalez
Ancud	ANC	-41.865314, -73.831927	South	Pyrifera	S. T. Gonzalez
Dalcahue	DAL	-42.382756, -73.655023	South	Pyrifera	S. T. Gonzalez
---

#### Molano et al. PRJNA661280
[Sporophyte Stage Genes Exhibit Stronger Selection Than Gametophyte Stage Genes in Haplodiplontic Giant Kelp](https://www.frontiersin.org/journals/marine-science/articles/10.3389/fmars.2021.774076/full)

---
Sampling locations	Longitude	Latitude
Catalina Island	-118.511	33.462
Carlsbad	-117.36	33.157
Santa Barbara/Arroyo Quemada	-120.119	34.47
---

#### Bemmels et al. PRJNA1140647/CCGP data

---
Site.ID      Lat      Long
AVI 35.24408 -120.9011
AVI 35.24124 -120.8955
BCP 33.44844 -118.4789
BCP 33.44768 -118.5117
BGC 36.06888 -121.6010
BOB 38.31320  123.0514
BOO 32.85220 -117.2760
CAR 36.26315 -121.8566
DAN 33.54213 -117.7950
HAL 33.52987 -117.7809
HAZ 35.28000 -120.8800
HAZ 35.28000 -120.8800
IRC 33.44768 -118.5117
LAJ 32.80959 -117.2866
MAL 34.02193 -118.7739
MILL 35.97976 -121.4905
MOH 34.39456 -119.7300
MPC 34.01923 -118.7877
PAO 33.70694 -118.2859
PV 33.72030 -118.3372
SHB 35.16902 -120.6971
SHB 35.16902 -120.6971
STL 36.56012 -121.9473
TER 36.94487 -122.0643
TWH 33.44549 -118.4983
TWH 33.42659 -118.5062
VEN 34.35552 -119.4412
---


#### Combine metadata and map

```{r, echo = FALSE, include = FALSE}
library(dplyr)
library(leaflet)
```

```{r, echo = FALSE}

# First dataset - Site codes with coordinates
alberto <- data.frame(
  Abbreviation = c("AM", "MO", "PL", "PP", "PV", "SC", "SM"),
  Site_Name = c("Andrew Molera", "Monterey", "Point Loma", "Santa Cruz",
                "Palos Verdes", "Santa Cruz Island", "San Miguel Island"),
  Latitude = c(36.283333, 36.617145, 32.678617, 36.952702, 33.720889, 34.032411, 34.051883),
  Longitude = c(-121.876392, -121.895693, -117.253367, -121.9711, -118.336278, -119.524502, -120.33885),
  Hemisphere = NA,
  Morph = NA,
  Collected_By = NA
)

# Second dataset - Main sampling data
gonzalez <- data.frame(
  Abbreviation = c("STL", "CAY", "AQ", "CI", "CB", "PLB", "PCH", "TOT",
                   "DOC", "ALG", "MON", "BOC", "TUM", "CHO", "BAM", "ANC", "DAL"),
  Site_Name = c("Stillwater Cove", "Cayucos", "Arroyo Quemado", "Catalina Island",
                "Camp Pendleton", "Playa Blanca", "Punta Choros", "Totorolillo",
                "Las Docas", "Algarrobo", "Las Monjas", "La Boca", "Tumbes",
                "Chome", "Bahía Mansa", "Ancud", "Dalcahue"),
  Latitude = c(36.565058, 35.446456, 34.47, 33.462, 33.157, -28.1851111,
               -29.245712, -32.022839, -33.139851, -33.360022, -33.492179,
               -33.903175, -36.623731, -36.806467, -40.581556, -41.865314, -42.382756),
  Longitude = c(-121.943553, -120.932394, -120.119, -118.511, -117.36, -71.1651667,
                -71.466174, -71.508658, -71.708218, -71.668728, -71.642485,
                -71.834772, -73.093019, -73.177088, -73.738197, -73.831927, -73.655023),
  Hemisphere = c("North", "North", "North", "North", "North", "South", "South",
                 "South", "South", "South", "South", "South", "South", "South",
                 "South", "South", "South"),
  Morph = c("Pyrifera and Integrifolia", "Pyrifera", "Pyrifera", "Pyrifera",
            "Pyrifera", "Integrifolia", "Integrifolia", "Integrifolia",
            "Integrifolia", "Integrifolia", "Integrifolia", "Pyrifera",
            "Pyrifera", "Pyrifera", "Pyrifera", "Pyrifera", "Pyrifera"),
  Collected_By = c("S. T. Gonzalez", "S. T. Gonzalez", "F. Alberto", "F. Alberto",
                   "F. Alberto", "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez",
                   "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez",
                   "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez",
                   "S. T. Gonzalez")
)

# Third dataset - Sampling locations
molano <- data.frame(
  Abbreviation = c("CI", "CB", "SB"),
  Site_Name = c("Catalina Island", "Carlsbad", "Santa Barbara/Arroyo Quemada"),
  Latitude = c(33.462, 33.157, 34.47),
  Longitude = c(-118.511, -117.36, -120.119),
  Hemisphere = NA,
  Morph = NA,
  Collected_By = NA
)

# CCGP
ccgp <- read.csv("./metadata/Mpyr_CCGP_ID.csv") |> distinct(Site.ID, Lat, Long, General.notes)
colnames(ccgp) <- c("site", "Latitude", "Longitude", "Notes")

# Combine all datasets
all_sites <- bind_rows(
  mutate(alberto, Dataset = "Alberto"),
  mutate(gonzalez, Dataset = "Gonzalez"),
  mutate(molano, Dataset = "Molano"), 
  mutate(ccgp, Dataset = "CCGP")
) %>%
  mutate(Row_ID = row_number())  # unique identifier for every row

# save 
write.table(all_sites, file = "./metadata/MpyrAdapt_sites.txt", row.names = FALSE, quote = FALSE, sep = ",")

# Color palette by Dataset
pal <- colorFactor(
  palette = c("#2196F3", "tomato", "#4CAF50", "#FF9800"),
  domain = c("Alberto", "Gonzalez", "Molano", "CCGP")
)

# Build the map
leaflet(all_sites) %>%
  addProviderTiles(providers$Esri.OceanBasemap) %>%
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    color = ~pal(Dataset),
    radius = 7,
    stroke = TRUE,
    weight = 1.5,
    fillOpacity = 0.85,
    popup = ~paste0(
      "<b>", Site_Name, "</b> (", Abbreviation, ")<br>",
      "Dataset: ", Dataset, "<br>",
      "Lat: ", round(Latitude, 4), ", Lon: ", round(Longitude, 4),
      ifelse(!is.na(Hemisphere), paste0("<br>Hemisphere: ", Hemisphere), ""),
      ifelse(!is.na(Morph), paste0("<br>Morph: ", Morph), ""),
      ifelse(!is.na(Collected_By), paste0("<br>Collected by: ", Collected_By), "")
    ),
    label = ~Site_Name
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~Dataset,
    title = "Dataset",
    opacity = 0.9
  ) %>%
  setView(lng = -100, lat = 0, zoom = 2)

```

# Data processing 

## Download genomes from SRA 

install SRA toolkit to download data from the SRA to elkhorn
```{bash, eval = FALSE}
# install SRA toolkit 
# Download the latest Linux binary
# from /home/miclark/software/
wget https://ftp-trace.ncbi.nlm.nih.gov/sra/sdk/current/sratoolkit.current-ubuntu64.tar.gz
tar -xzf sratoolkit.current-ubuntu64.tar.gz

# Add to PATH
echo 'export PATH=$HOME/software/sratoolkit.3.3.0-ubuntu64/bin:$PATH' >> ~/.bashrc
source ~/.bashrc

# configure SRA toolkit
# follow guidance from: https://github.com/ncbi/sra-tools/wiki/05.-Toolkit-Configuration
vdb-config --interactive
  # in cache, set user-repository to /hb/scratch/miclark/sra

```

Download SRA data
```{bash, eval = FALSE}
sbatch ./scripts/sra_download.sbatch PRJNA938791 # Submitted batch job 106781
    # "SRR023" is text from the URL that hosts the sequences, included in SRR_accessions accidently 
sbatch ./scripts/sra_download.sbatch PRJNA661280 # Submitted batch job 106787
    # "SRR012" is text from the URL that hosts the sequences, included in SRR_accessions accidently 

# compression is taking a long time -- make parallel in an interactive job
  # cancelling when compressing SRR23817627_1.fastq and SRR12582529_2.fastq 
salloc --time=24:00:00 --mem=100Gb --ntasks=1 --cpus-per-task=32 --partition=lab-mpinsky \
        --qos=pi-mpinsky \
        --account=pi-mpinsky

# PRJNA661280
INDIR=/home/miclark/Mpyr_WGS_data/PRJNA661280/fastq
module load parallel/20200122
ls ${INDIR}/*.fastq | parallel --no-notice -j 32 "gzip {}"
rm /home/miclark/Mpyr_WGS_data/PRJNA661280/fastq/SRR12582529_2.fastq.gz
gzip /home/miclark/Mpyr_WGS_data/PRJNA661280/fastq/SRR12582529_2.fastq

# PRJNA938791 [x]
INDIR=/home/miclark/Mpyr_WGS_data/PRJNA938791/fastq
rm /home/miclark/Mpyr_WGS_data/PRJNA938791/fastq/SRR23817627_1.fastq.gz # [x] 
ls ${INDIR}/*.fastq | parallel --no-notice -j 32 "gzip {}" # [x]

# moved sra files to scratch 

# generate md5sums 
cd /home/miclark/Mpyr_WGS_data/PRJNA938791/fastq/
md5sum ./*.fastq.gz > ./md5sums.txt
cd /home/miclark/Mpyr_WGS_data/PRJNA661280/fastq/
md5sum ./*.fastq.gz > ./md5sums.txt
cd /home/miclark/Mpyr_WGS_data/Mpyr_Alberto_data
md5sum ./*.fastq.gz > ./md5sums.txt

# move fastq.gz files to scratch 
mv /home/miclark/Mpyr_WGS_data/* /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/ #[x]

# verify md5sums
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/fastq/
md5sum -c ./md5sums.txt > ./md5sums_check.txt # [x]
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA661280/fastq/
md5sum -c ./md5sums.txt > ./md5sums_check.txt # [x]
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/Mpyr_Alberto_data
md5sum -c ./md5sums.txt > ./md5sums_check.txt # []

# get run information, which includes individual sample names, saves to runinfo.csv in each project directory
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/
wget -q "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=sra&id=$(grep -o '"[0-9]\{6,\}"' ./search.json | tr -d '"' | paste -sd',')&rettype=runinfo&retmode=csv" -O runinfo.csv
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA661280/
wget -q "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=sra&id=$(grep -o '"[0-9]\{6,\}"' ./search.json | tr -d '"' | paste -sd',')&rettype=runinfo&retmode=csv" -O runinfo.csv


```

## Raw data processing 

Standardize file names and set up symbolic links from Mpyr_WGS_data to MpyrAdapt/data/raw \
  - Naming convetion: Species_project_site-sample_number_readDirection, e.g. Mpyr_ALB_STL-03_1, Mpyr_MOL_STL-03_2, Mpyr_GON_STL-03_1

```{bash, eval = FALSE}

# set up symbolic links from Mpyr_WGS_data to MpyrAdapt/data/raw
sbatch ./scripts/make_symbolic_links.sbatch # Submitted batch job 109551
sbatch ./scripts/make_symbolic_links_CCGP.sbatch # Submitted batch job 155406

# two files not found
    # WARNING: /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/fastq/SRR23817635_1.fastq.gz not found, skipping
    # WARNING: /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/fastq/SRR23817635_2.fastq.gz not found, skipping
    # SRR23817635 is a sequence that exists from Gonzalez et al, but is missing from sra download. This is a southern hemisphere integrifolia sample, which is likely outside the scope of this project, so we will proceed without it 

# set up symbolic link to reference genome
ln -s /hb/groups/mpinsky_lab/Mpyr_ref_2025/hap1/Mpyr-NLJ1B.v3.hap1.softmasked.fasta /home/miclark/MpyrAdapt/data/reference/Mpyr-NLJ1B.v3.hap1.softmasked.fasta

```

Run FastQC on raw reads, run MultiQC to summarize results, and check percent adapter dimer
```{bash, eval = FALSE}
sbatch ./scripts/read_QC.sbatch "/home/miclark/MpyrAdapt/data/raw" "raw_reads_QC" # Submitted batch job 109557, Submitted batch job 155663

```

Trim raw reads with Fastp
```{bash, eval = FALSE}
sbatch ./scripts/trim_reads.sbatch

# job finished fastp, then stopped running before running multiQC. Not sure if this is an issue with the node, with the script, or with the mutliQC module

# running multiQC manually: 
#run multiqc 
salloc --time=24:00:00 --mem=50Gb --ntasks=1 --cpus-per-task=1 --partition=lab-mpinsky \
        --qos=pi-mpinsky \
        --account=pi-mpinsky

OUTDIR=/home/miclark/MpyrAdapt/data/processed

module load multiqc

multiqc -v -p -ip -f --data-dir --data-format tsv --cl-config "max_table_rows: 3000" --filename trimmed_report --outdir $OUTDIR/QC $OUTDIR

# ran in < 2 minutes! 

```

## Align to reference genome

Make array key for alignment to reference genome
```{bash, eval = FALSE}
# code in make_alignment_key.sbatch
# script to make alignment array key to run with wrapper-align_to_genome.sh

KEY=/home/miclark/MpyrAdapt/scripts/alignment_key.txt
INDIR=/home/miclark/MpyrAdapt/data/processed/merged

ls $INDIR/*mrg.fq.gz | \
    sed -E 's/.*\///' | \
    sed -E 's/_mrg\.fq\.gz$//' | \
    while read line; do
        echo -e "$line\t$INDIR/${line}_r1.fq.gz\t$INDIR/${line}_r2.fq.gz\t$INDIR/${line}_mrg.fq.gz"
    done > $KEY

```

Install bwa on hummingbird––using bwa-mem2 for now, but bwa is installed in case I need to use aln in the future
```{bash, eval = FALSE}
module load miniconda3
conda create -n bwa
conda activate bwa 
conda install -c bioconda -c conda-forge bwa

```

Index reference genome with bwa
```{bash, eval = FALSE}
module load bwa-mem2/2.2.1
reference=/home/miclark/MpyrAdapt/data/reference/Mpyr-NLJ1B.v3.hap1.softmasked.fasta
bwa-mem2 index $reference

```

Start array job to align to genome
```{bash, eval = FALSE}
sbatch ./scripts/wrapper-align_to_genome.sh # submitted as array job 126960

# just do mem on merged and unmerged to begin with, if things look crazy, can compare mem and aln 

# I was taking up the entire reserach partition, so I cancelled arrays 68-188, will re-run just those arrays with a node limit. 
  # I also altered align_to_genome.sbatch so that it doesn't have to decompress read files multiple times to speed up jobs
  # submitted as array job 151608 and 151660

# running separate alignment job with CCGP_alignment_key.txt for CCGP data, submitted as array job 181599
```

Install qualimap
```{bash, eval = FALSE}
module load miniconda3

conda create -n qualimap -c bioconda -c conda-forge qualimap
conda activate qualimap

qualimap --help
```

Run qualimap on aligned BAM files to check mapping quality and coverage
```{bash, eval = FALSE}
./scripts/wrapper-run_qualimap.sh # jobid 152761
# running on CCGP data separately, job id 181803

# summarize qualimap output with multiQC 
sbatch ./scripts/summarize_qualimap.sbatch # jobid 154531, 182483 with CCGP data 
```

Install bamstats 
```{bash, eval = FALSE}
# in /home/miclark/software directory: 
module load git
git clone https://github.com/tplinderoth/ngsQC.git

cd /ngsQC/bamstats
g++ -O3 -std=c++11 -o bamstats bamstats.cpp

```

Merge bam files 
```{bash, eval = FALSE} 
# one single bam is required for bamstats

sbatch merge_bams.sbatch # jobid 198203

# Mega bam file is quite large and stored on scratch 
``` 

Run bamstats for each scaffold
```{bash, eval = FALSE}

./scripts/wrapper-run_bamstats.sh # job id 199063

``` 

Summarize bamstats output as histograms 
```{bash, eval = FALSE}
# Read output files into R and plot coverage histograms to inform depth cut-offs for variant calling

sbatch ./scripts/make_bamQC_histograms.sbatch # job id 199350

```

Based on bamstats output, reasonable depth cut-offs are: 
- min depth: 871 reads (median * 0.5) 
- max depth: 3486 reads (median * 2) 

*OUTDATED*: Use samtools depth to generate per-site coverage histogram to inform depth cut-offs
```{bash, eval = FALSE}
# Strategy: job will be an array job, where each task is a different chromosome, so the list of chromosomes will act as a array key. Each task will use all bams

# index .bam files (only need to happen once) from an interactive job
salloc --time=24:00:00 --mem=100Gb --ntasks=1 --cpus-per-task=10 --partition=lab-mpinsky \
        --qos=pi-mpinsky \
        --account=pi-mpinsky
module load samtools
indir="/home/miclark/MpyrAdapt/data/alignments"
samtools index -M -@ 10 ${indir}/*.bam 

# chrom_list.txt is a list of chromosome names to use as an array key for the samtools depth job, which will run one chromosome per task in an array job

./scripts/wrapper-bam_depth.sh 

# Read output files into R and plot coverage histograms to inform depth cut-offs for variant calling

sbatch ./scripts/make_depth_histograms.sbatch # jobid 198073

```

## Make pass/fail masks 
Make pass/fail .bed files based on depth and mapping quality cut-offs
```{bash, eval = FALSE}

./scripts/wrapper-generate_PF_masks.sh # job id 199424

# summarize pass/fail reports
reportdir=/home/miclark/MpyrAdapt/data/variants/masks/reports
for f in ${reportdir}/*.txt; do
  awk '/Number sites analyzed:/{total=$NF} /Number FAIL sites:/{fail=$NF} END{printf "%s: %.2f%%\n", FILENAME, (fail/total)*100}' "$f"
done

```
Based on pass/fail reports, ~20-25% of sites genome wide were marked as FAIL for most scaffolds. Scaffols 31 was almost entirely FAILed 

---
MpyrAdapt_PF_mask_10.txt: 30.59%
MpyrAdapt_PF_mask_11.txt: 22.27%
MpyrAdapt_PF_mask_12.txt: 46.66%
MpyrAdapt_PF_mask_13.txt: 33.28%
MpyrAdapt_PF_mask_14.txt: 28.12%
MpyrAdapt_PF_mask_15.txt: 16.21%
MpyrAdapt_PF_mask_16.txt: 29.15%
MpyrAdapt_PF_mask_17.txt: 27.16%
MpyrAdapt_PF_mask_18.txt: 34.79%
MpyrAdapt_PF_mask_19.txt: 23.64%
MpyrAdapt_PF_mask_1.txt: 24.81%
MpyrAdapt_PF_mask_20.txt: 31.22%
MpyrAdapt_PF_mask_21.txt: 20.02%
MpyrAdapt_PF_mask_22.txt: 21.31%
MpyrAdapt_PF_mask_23.txt: 33.54%
MpyrAdapt_PF_mask_24.txt: 46.26%
MpyrAdapt_PF_mask_25.txt: 16.66%
MpyrAdapt_PF_mask_26.txt: 26.16%
MpyrAdapt_PF_mask_27.txt: 19.86%
MpyrAdapt_PF_mask_28.txt: 17.77%
MpyrAdapt_PF_mask_29.txt: 20.46%
MpyrAdapt_PF_mask_2.txt: 29.84%
MpyrAdapt_PF_mask_30.txt: 18.12%
MpyrAdapt_PF_mask_31.txt: 100.00%
MpyrAdapt_PF_mask_3.txt: 23.04%
MpyrAdapt_PF_mask_4.txt: 20.66%
MpyrAdapt_PF_mask_5.txt: 29.51%
MpyrAdapt_PF_mask_6.txt: 24.03%
MpyrAdapt_PF_mask_7.txt: 22.96%
MpyrAdapt_PF_mask_8.txt: 24.18%
MpyrAdapt_PF_mask_9.txt: 17.60%
---


## Call SNPs
For this project, we will use a join genotype call and genotype liklihood approach. 
- Genotypes will be called with samples with >10X mean depth
- Genotype liklihoods will be calucated for all samples 
- Analyses with GLs can be confirmed with genotype calls 
- Approach: call SNPs with bcftools mpileup and bcftools call. Loci will be filtered in two flavors: one for liklihoods and one for calls 

Use bcftools mpileup and call to call SNPs and calculate genotype likelihoods across all samples
```{bash, eval = FALSE}

./scripts/wrapper-call_snps.sh # jobid 194785

```

Normalize variants
```{bash, eval = FALSE}

./scripts/wrapper-normalize_snps.sh # job id 199094

```

## Annotate VCF

Make group file
```{bash, eval = FALSE}
indir="/hb/scratch/miclark/MpyrAdapt/variants/normBCFs"
outdir="/home/miclark/MpyrAdapt/scripts/"

# get list of individuals in bcf file
bcftools view ${indir}/MpyrAdapt_scaffold_31_norm.bcf.gz | grep -m1 "^#CHROM" | cut -f 10- | tr "\t" "\n" > $outdir/all_individuals.txt

module load r
```

```{r, eval = FALSE

# read in list of individuals generated above
inds <- read.csv("/home/miclark/MpyrAdapt/scripts/all_individuals.txt", header = FALSE)$V1

# get sites of origin, clean naming differences (whoops) 
sites_of_origin <- unlist(lapply(X = strsplit(inds, "_"), FUN = function(x){return(x[3])}))
sites_of_origin <- unlist(lapply(X = strsplit(sites_of_origin, "-"), FUN = function(x){return(x[1])}))

sites <- unique(sites_of_origin) $ 42 sites! 

# create groups file: 

# column 1: #GROUP
sites  

# column 2: INDEX
index_vec <- vector(length = length(sites))
for(i in 1:length(sites)){
  index_vec[i] <- paste(which(sites_of_origin == sites[i]) - 1, sep="", collapse=",")
  print(paste0(i, " ", round(0.9*length(which(sites_of_origin == sites[i])))))
}

grp_data <- cbind.data.frame(sites, index_vec, rep(15, length = length(sites)), rep(5, length = length(sites)))
colnames(grp_data) <- c("#GROUP", "INDEX", "MIN_GQ", "MIN_DP")
write.table(grp_data, file = "/home/miclark/MpyrAdapt/scripts/Mpyr_groups_qc.txt", 
            quote = FALSE, sep = "\t", row.names=FALSE)


```


Run insertAnnotations_Mpyr.pl
```{bash, eval = FALSE}

./wrapper-annotate_VCF.sh # job id 199582

  # needed to replace MoreUtils perl module with Util, probably because elkhorn has a newer version of perl than the script was written for. 
```


## Filter VCF 
To get initial sense of batch effects and initial analyses, I will just filtering SNPs, ignoring nonvariant sites. 
My strategy is to get a filter of high quality SNP sites. Then, those sites will be used to get genotype likelihoods. 
If many individuals fail, I may want to go back and exclude them from bamstats when deciding cut-offs. 
*Eventually, I probably want to filter individuals from coverage at the bam stage* 

Get coverage at PASS sites to determine if I should drop any low coverage individuals 
```{bash, eval = FALSE}

sbatch ./scripts/calc_depth_vcf.sbatch # job id 200274

```


Filter out FAIL individuals 
```{bash, eval = FALSE} 


# Output: list of individuals that pass quality filtering to use as BAMLIST when converting genotype likelihoods 

``` 


Make line of code to filter SNPs
```{r, eval = FALSE}

# read in list of individuals generated above
inds <- read.csv("/home/miclark/MpyrAdapt/scripts/all_individuals.txt", header = FALSE)$V1 # update for high quality individuals 

# get sites of origin, clean naming differences (whoops) 
sites_of_origin <- unlist(lapply(X = strsplit(inds, "_"), FUN = function(x){return(x[3])}))
sites_of_origin <- unlist(lapply(X = strsplit(sites_of_origin, "-"), FUN = function(x){return(x[1])}))

sites <- unique(sites_of_origin) # 42 sites! 

index_vec <- vector(length = length(sites))
for(i in 1:length(sites)){
  index_vec[i] <- paste(which(sites_of_origin == sites[i]) - 1, sep="", collapse=",")
  print(paste0(i, " ", round(0.9*length(which(sites_of_origin == sites[i])))))
}

grp_data <- cbind.data.frame(sites, index_vec, rep(15, length = length(sites)), rep(2, length = length(sites)))
colnames(grp_data) <- c("#GROUP", "INDEX", "MIN_GQ", "MIN_DP")

# make filtering line 

### depth part
for(i in 1:nrow(grp_data)){
  cat(" N_PASS(GT[", grp_data[i,"INDEX"], "]!=\"mis\" & FMT/DP >= 2) >= ", round(length(unlist(strsplit(grp_data[i,"INDEX"], ",")))*.9), " &&", 
      sep = "")
}

### genotype quality part: 
for(i in 1:nrow(grp_data)){
  cat("$", i, " >= ", round(length(unlist(strsplit(grp_data[i,"INDEX"], ",")))*.9), " && ",
      sep = "")
}


```


Make .pos file of sites that pass preliminary filtering to pass to angsd
```{bash, eval = FALSE}

# extract_qc_SNPs.sbatch 
# use list of individuals generated above for filtering
# make per scaffold using a wrapper, then merge pos files together

#!/bin/bash

########## SBATCH Lines for Resource Request ##########

#SBATCH --time=48:00:00             # limit of wall clock time - how long the job will run (same as -t)
#SBATCH --cpus-per-task=1         # number of CPUs (or cores) per task (same as -c)
#SBATCH --mem-per-cpu=10G            # memory required per allocated CPU (or core)
#SBATCH --job-name=extract_qc_pos      # you can give your job a name for easier identification (same as -J)
#SBATCH --output="/mnt/home/clarkm89/petrels/log_extract_qc_pos/extract_qc_snps_%A.out"
#SBATCH --error="/mnt/home/clarkm89/petrels/log_extract_qc_pos/extract_qc_snps_%A.err"
#SBATCH --account=bradburd

##########

VCF="/mnt/home/clarkm89/petrels/called_snps/annotated_vcf/${SPP}_norm_annotated.vcf.gz"
SITES="/mnt/home/clarkm89/petrels/called_snps/annotated_vcf/${SPP}_qc_snps.pos"
AB_CONTIGS="/mnt/home/clarkm89/petrels/allelic_balance/${SPP}_exclude_regions.txt"

# 90% of individuals in a population need to have at least two reads 
# filter low coverage individual 
# bcftools view -s "/mnt/home/clarkm89/petrels/alignments/ref/MARK_322.rmdup.bam,/mnt/home/clarkm89/petrels/alignments/ref/MARK_329.rmdup.bam,/mnt/home/clarkm89/petrels/alignments/ref/MASP_13.rmdup.bam,/mnt/home/clarkm89/petrels/alignments/ref/MASP_16.rmdup.bam,/mnt/home/clarkm89/petrels/alignments/ref/MASP_7.rmdup.bam,/mnt/home/clarkm89/petrels/alignments/ref/MASP_10.rmdup.bam" $VCF \
#| bcftools view -f "PASS" -i 'N_PASS(FMT/DP[0-1] > 1) > 1 && N_PASS(FMT/DP[2-5] > 1) > 3' -m 2 -M 2 -v snps \
#| bcftools query -f '%CHROM\t%POS\n' | uniq > $SITES
bcftools view -f "PASS" -T "^${AB_CONTIGS}" -i 'N_PASS(FMT/DP[0-1] > 1) > 1 && N_PASS(FMT/DP[2-5] > 1) > 3' -m 2 -M 2 -v snps $VCF \
  bcftools query -f '%CHROM\t%POS\n' | uniq > $SITES

wait

angsd sites index $SITES

printf "\nfinished\n"



```


## Convert to genotype likelihoods
requires: 
- list of qc-passed and LD-pruned SNPs where we trust genotype likelihoods. Maybe skip LD prune for now? 
- 
```{bash, eval = FALSE}

#!/bin/bash

# Script written by M. Clark, 2/24/2023
# This script calculates genotype likelihoods from a list of bamfiles

# usage: 
# call_geno_likelihood.sh

module list

SPP=$1
OUTDIR=${SPP}_geno_likelihoods_LD
SITES=/mnt/home/clarkm89/petrels/called_snps/annotated_vcf/LD_prune/${SPP}_norm_annotated_qcLDPrune.pos
BAMLIST=/mnt/home/clarkm89/petrels/${SPP}_bamlist.txt

if [ ! -d ./$OUTDIR ]; then mkdir ./$OUTDIR; fi

# trim pos file 

cut -f1 ${SITES} | sort | uniq > ${OUTDIR}/${SPP}_chrs.txt

# index pos file
angsd sites index ${SITES}

# index bam files 
while read bam; do
  if [ ! -f "${bam}.bai" ]; then
    echo "Indexing $bam..."
    samtools index "$bam"
  else
    echo "Index exists for $bam, skipping..."
  fi
done < $BAMLIST

angsd -b ${BAMLIST} -GL 1 -doMaf 1 -doMajorMinor 1 -SNP_pval 1e-4 -doGlf 2 -sites ${SITES} -rf ${OUTDIR}/${SPP}_chrs.txt -out ./$OUTDIR/${SPP}_gl

echo done calling genotype likelihoods


```

# Analyses 

## PCA 


## Heterozygosity, pi, theta


## Inbreeding coefficients 
- Correlate with heatwave intensity? 


## Genotype-environment association
- Basic GEA with RDA? Bishop et al. uses "Capblancq and Forester (2021) using the R package ‘vegan’ version 2.6.4."


