---
title: "Macrocystis WGS Project"
author: "Meaghan Clark"
date: "2026-02-24"
output: html_document
editor_options: 
  chunk_output_type: console
---
## Summary 

Project begun Februray 2026. Goal is to combine multiple *Macrocystis* whole genome resequencing data sets to learn about structure, inbreeding, and local adaptation. 


## Directory structure

```
MpyrAdapt/
├── data/
│   ├── raw/          # symbolic links to raw sequencing data
│   ├── processed/    # trimmed, filtered, aligned data
│   └── reference/    # symbolic link to reference genome
├── scripts/
│   ├── 0. sra_download.sbatch
│   ├── 1. make_smbolic_links.sbatch
│   ├── 1.5 make_smbolic_links_CCGP.sbatch
│   ├── 2. trim_reads.sbatch
│   ├── 3. make_alignment_key.sbatch
│   ├── 4. wrapper-align_to_genome.sh
│   ├──── align_to_genome.sbatch
│   ├── 5. wrapper-run_qualimap.sh
│   ├──── run_qualimap.sbatch
│   ├── 6. summarize_qualimap.sbatch
│   ├── 7. wrapper-bam_depth.sh
│   ├──── bam_depth.sbatch
│   ├── 8. make_depth_histograms.R
│   ├── 9. wrapper-call_snps.sh
│   ├──── call_snps.sbatch
│   ├── 
│   ├── 
│   ├── 
│   ├── 
│   ├── 
│   ├── 

│   └── 
├── 
│   ├── 
│   └── 
├── logs
└── MpyrAdapt_notebook.md
```

## Metadata 

##### Site keys from Alberto data 

---
AM	Andrew Molera	36.283333	-121.876392
MO	Monterey	36.617145	-121.895693
PL	Point Loma	32.678617	-117.253367
PP	Santa Cruz	36.952702	-121.9711
PV	Palos Verdes	33.720889	-118.336278
SC	Santa Cruz Island	34.032411	-119.524502
SM	San Miguel Island	34.051883	-120.33885
---

#### Gonzalez et al. PRJNA938791  
[Whole-genome sequencing distinguishes the two most common giant kelp ecomorphs](https://academic.oup.com/evolut/article/77/6/1354/7079449)

---
Site name	Site abbreviation	Coordinates	Hemisphere	Morph	Collected by
Stillwater Cove	STL	36.565058, -121.943553	North	Pyrifera and Integrifolia	S. T. Gonzalez
Cayucos	CAY	35.446456, -120.932394	North	Pyrifera	S. T. Gonzalez
Arroyo Quemado	AQ	34.47, -120.119	North	Pyrifera	F. Alberto
Catalina Island	CI	33.462, -118.511	North	Pyrifera	F. Alberto
Camp Pendleton	CB	33.157, -117.36	North	Pyrifera	F. Alberto
Playa Blanca	PLB	-28.1851111, -71.1651667	South	Integrifolia	S. T. Gonzalez
Punta Choros	PCH	-29.245712, -71.466174	South	Integrifolia	S. T. Gonzalez
Totorolillo	TOT	-32.022839, -71.508658	South	Integrifolia	S. T. Gonzalez
Las Docas	DOC	-33.139851, -71.708218	South	Integrifolia	S. T. Gonzalez
Algarrobo	ALG	-33.360022, -71.668728	South	Integrifolia	S. T. Gonzalez
Las Monjas	MON	-33.492179, -71.642485	South	Integrifolia	S. T. Gonzalez
La Boca	BOC	-33.903175, -71.834772	South	Pyrifera	S. T. Gonzalez
Tumbes	TUM	-36.623731, -73.093019	South	Pyrifera	S. T. Gonzalez
Chome	CHO	-36.806467, -73.177088	South	Pyrifera	S. T. Gonzalez
Bahía Mansa	BAM	-40.581556, -73.738197	South	Pyrifera	S. T. Gonzalez
Ancud	ANC	-41.865314, -73.831927	South	Pyrifera	S. T. Gonzalez
Dalcahue	DAL	-42.382756, -73.655023	South	Pyrifera	S. T. Gonzalez
---

#### Molano et al. PRJNA661280
[Sporophyte Stage Genes Exhibit Stronger Selection Than Gametophyte Stage Genes in Haplodiplontic Giant Kelp](https://www.frontiersin.org/journals/marine-science/articles/10.3389/fmars.2021.774076/full)

---
Sampling locations	Longitude	Latitude
Catalina Island	-118.511	33.462
Carlsbad	-117.36	33.157
Santa Barbara/Arroyo Quemada	-120.119	34.47
---

#### Bemmels et al. PRJNA1140647/CCGP data

---
Site.ID      Lat      Long
AVI 35.24408 -120.9011
AVI 35.24124 -120.8955
BCP 33.44844 -118.4789
BCP 33.44768 -118.5117
BGC 36.06888 -121.6010
BOB 38.31320  123.0514
BOO 32.85220 -117.2760
CAR 36.26315 -121.8566
DAN 33.54213 -117.7950
HAL 33.52987 -117.7809
HAZ 35.28000 -120.8800
HAZ 35.28000 -120.8800
IRC 33.44768 -118.5117
LAJ 32.80959 -117.2866
MAL 34.02193 -118.7739
MILL 35.97976 -121.4905
MOH 34.39456 -119.7300
MPC 34.01923 -118.7877
PAO 33.70694 -118.2859
PV 33.72030 -118.3372
SHB 35.16902 -120.6971
SHB 35.16902 -120.6971
STL 36.56012 -121.9473
TER 36.94487 -122.0643
TWH 33.44549 -118.4983
TWH 33.42659 -118.5062
VEN 34.35552 -119.4412
---


#### Combine metadata and map

```{r, echo = FALSE, include = FALSE}
library(dplyr)
library(leaflet)
```

```{r, echo = FALSE}

# First dataset - Site codes with coordinates
alberto <- data.frame(
  Abbreviation = c("AM", "MO", "PL", "PP", "PV", "SC", "SM"),
  Site_Name = c("Andrew Molera", "Monterey", "Point Loma", "Santa Cruz",
                "Palos Verdes", "Santa Cruz Island", "San Miguel Island"),
  Latitude = c(36.283333, 36.617145, 32.678617, 36.952702, 33.720889, 34.032411, 34.051883),
  Longitude = c(-121.876392, -121.895693, -117.253367, -121.9711, -118.336278, -119.524502, -120.33885),
  Hemisphere = NA,
  Morph = NA,
  Collected_By = NA
)

# Second dataset - Main sampling data
gonzalez <- data.frame(
  Abbreviation = c("STL", "CAY", "AQ", "CI", "CB", "PLB", "PCH", "TOT",
                   "DOC", "ALG", "MON", "BOC", "TUM", "CHO", "BAM", "ANC", "DAL"),
  Site_Name = c("Stillwater Cove", "Cayucos", "Arroyo Quemado", "Catalina Island",
                "Camp Pendleton", "Playa Blanca", "Punta Choros", "Totorolillo",
                "Las Docas", "Algarrobo", "Las Monjas", "La Boca", "Tumbes",
                "Chome", "Bahía Mansa", "Ancud", "Dalcahue"),
  Latitude = c(36.565058, 35.446456, 34.47, 33.462, 33.157, -28.1851111,
               -29.245712, -32.022839, -33.139851, -33.360022, -33.492179,
               -33.903175, -36.623731, -36.806467, -40.581556, -41.865314, -42.382756),
  Longitude = c(-121.943553, -120.932394, -120.119, -118.511, -117.36, -71.1651667,
                -71.466174, -71.508658, -71.708218, -71.668728, -71.642485,
                -71.834772, -73.093019, -73.177088, -73.738197, -73.831927, -73.655023),
  Hemisphere = c("North", "North", "North", "North", "North", "South", "South",
                 "South", "South", "South", "South", "South", "South", "South",
                 "South", "South", "South"),
  Morph = c("Pyrifera and Integrifolia", "Pyrifera", "Pyrifera", "Pyrifera",
            "Pyrifera", "Integrifolia", "Integrifolia", "Integrifolia",
            "Integrifolia", "Integrifolia", "Integrifolia", "Pyrifera",
            "Pyrifera", "Pyrifera", "Pyrifera", "Pyrifera", "Pyrifera"),
  Collected_By = c("S. T. Gonzalez", "S. T. Gonzalez", "F. Alberto", "F. Alberto",
                   "F. Alberto", "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez",
                   "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez",
                   "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez", "S. T. Gonzalez",
                   "S. T. Gonzalez")
)

# Third dataset - Sampling locations
molano <- data.frame(
  Abbreviation = c("CI", "CB", "SB"),
  Site_Name = c("Catalina Island", "Carlsbad", "Santa Barbara/Arroyo Quemada"),
  Latitude = c(33.462, 33.157, 34.47),
  Longitude = c(-118.511, -117.36, -120.119),
  Hemisphere = NA,
  Morph = NA,
  Collected_By = NA
)

# CCGP
ccgp <- read.csv("./metadata/Mpyr_CCGP_ID.csv") |> distinct(Site.ID, Lat, Long, General.notes)
colnames(ccgp) <- c("site", "Latitude", "Longitude", "Notes")

# Combine all datasets
all_sites <- bind_rows(
  mutate(alberto, Dataset = "Alberto"),
  mutate(gonzalez, Dataset = "Gonzalez"),
  mutate(molano, Dataset = "Molano"), 
  mutate(ccgp, Dataset = "CCGP")
) %>%
  mutate(Row_ID = row_number())  # unique identifier for every row

# Color palette by Dataset
pal <- colorFactor(
  palette = c("#2196F3", "tomato", "#4CAF50", "#FF9800"),
  domain = c("Alberto", "Gonzalez", "Molano", "CCGP")
)

# Build the map
leaflet(all_sites) %>%
  addProviderTiles(providers$Esri.OceanBasemap) %>%
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    color = ~pal(Dataset),
    radius = 7,
    stroke = TRUE,
    weight = 1.5,
    fillOpacity = 0.85,
    popup = ~paste0(
      "<b>", Site_Name, "</b> (", Abbreviation, ")<br>",
      "Dataset: ", Dataset, "<br>",
      "Lat: ", round(Latitude, 4), ", Lon: ", round(Longitude, 4),
      ifelse(!is.na(Hemisphere), paste0("<br>Hemisphere: ", Hemisphere), ""),
      ifelse(!is.na(Morph), paste0("<br>Morph: ", Morph), ""),
      ifelse(!is.na(Collected_By), paste0("<br>Collected by: ", Collected_By), "")
    ),
    label = ~Site_Name
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~Dataset,
    title = "Dataset",
    opacity = 0.9
  ) %>%
  setView(lng = -100, lat = 0, zoom = 2)

```

#### Download genomes from SRA 

install SRA toolkit to download data from the SRA to elkhorn
```{bash, eval = FALSE}
# install SRA toolkit 
# Download the latest Linux binary
# from /home/miclark/software/
wget https://ftp-trace.ncbi.nlm.nih.gov/sra/sdk/current/sratoolkit.current-ubuntu64.tar.gz
tar -xzf sratoolkit.current-ubuntu64.tar.gz

# Add to PATH
echo 'export PATH=$HOME/software/sratoolkit.3.3.0-ubuntu64/bin:$PATH' >> ~/.bashrc
source ~/.bashrc

# configure SRA toolkit
# follow guidance from: https://github.com/ncbi/sra-tools/wiki/05.-Toolkit-Configuration
vdb-config --interactive
  # in cache, set user-repository to /hb/scratch/miclark/sra

```

Download SRA data
```{bash, eval = FALSE}
sbatch ./scripts/sra_download.sbatch PRJNA938791 # Submitted batch job 106781
    # "SRR023" is text from the URL that hosts the sequences, included in SRR_accessions accidently 
sbatch ./scripts/sra_download.sbatch PRJNA661280 # Submitted batch job 106787
    # "SRR012" is text from the URL that hosts the sequences, included in SRR_accessions accidently 

# compression is taking a long time -- make parallel in an interactive job
  # cancelling when compressing SRR23817627_1.fastq and SRR12582529_2.fastq 
salloc --time=24:00:00 --mem=100Gb --ntasks=1 --cpus-per-task=32 --partition=lab-mpinsky \
        --qos=pi-mpinsky \
        --account=pi-mpinsky

# PRJNA661280
INDIR=/home/miclark/Mpyr_WGS_data/PRJNA661280/fastq
module load parallel/20200122
ls ${INDIR}/*.fastq | parallel --no-notice -j 32 "gzip {}"
rm /home/miclark/Mpyr_WGS_data/PRJNA661280/fastq/SRR12582529_2.fastq.gz
gzip /home/miclark/Mpyr_WGS_data/PRJNA661280/fastq/SRR12582529_2.fastq

# PRJNA938791 [x]
INDIR=/home/miclark/Mpyr_WGS_data/PRJNA938791/fastq
rm /home/miclark/Mpyr_WGS_data/PRJNA938791/fastq/SRR23817627_1.fastq.gz # [x] 
ls ${INDIR}/*.fastq | parallel --no-notice -j 32 "gzip {}" # [x]

# moved sra files to scratch 

# generate md5sums 
cd /home/miclark/Mpyr_WGS_data/PRJNA938791/fastq/
md5sum ./*.fastq.gz > ./md5sums.txt
cd /home/miclark/Mpyr_WGS_data/PRJNA661280/fastq/
md5sum ./*.fastq.gz > ./md5sums.txt
cd /home/miclark/Mpyr_WGS_data/Mpyr_Alberto_data
md5sum ./*.fastq.gz > ./md5sums.txt

# move fastq.gz files to scratch 
mv /home/miclark/Mpyr_WGS_data/* /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/ #[x]

# verify md5sums
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/fastq/
md5sum -c ./md5sums.txt > ./md5sums_check.txt # [x]
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA661280/fastq/
md5sum -c ./md5sums.txt > ./md5sums_check.txt # [x]
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/Mpyr_Alberto_data
md5sum -c ./md5sums.txt > ./md5sums_check.txt # []

# get run information, which includes individual sample names, saves to runinfo.csv in each project directory
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/
wget -q "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=sra&id=$(grep -o '"[0-9]\{6,\}"' ./search.json | tr -d '"' | paste -sd',')&rettype=runinfo&retmode=csv" -O runinfo.csv
cd /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA661280/
wget -q "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=sra&id=$(grep -o '"[0-9]\{6,\}"' ./search.json | tr -d '"' | paste -sd',')&rettype=runinfo&retmode=csv" -O runinfo.csv


```

## Raw data processing 

Standardize file names and set up symbolic links from Mpyr_WGS_data to MpyrAdapt/data/raw \
  - Naming convetion: Species_project_site-sample_number_readDirection, e.g. Mpyr_ALB_STL-03_1, Mpyr_MOL_STL-03_2, Mpyr_GON_STL-03_1

```{bash, eval = FALSE}

# set up symbolic links from Mpyr_WGS_data to MpyrAdapt/data/raw
sbatch ./scripts/make_symbolic_links.sbatch # Submitted batch job 109551
sbatch ./scripts/make_symbolic_links_CCGP.sbatch # Submitted batch job 155406

# two files not found
    # WARNING: /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/fastq/SRR23817635_1.fastq.gz not found, skipping
    # WARNING: /data/rufus/pinsky_lab/Mpyr/non_GCRG_WGS/PRJNA938791/fastq/SRR23817635_2.fastq.gz not found, skipping
    # SRR23817635 is a sequence that exists from Gonzalez et al, but is missing from sra download. This is a southern hemisphere integrifolia sample, which is likely outside the scope of this project, so we will proceed without it 

# set up symbolic link to reference genome
ln -s /hb/groups/mpinsky_lab/Mpyr_ref_2025/hap1/Mpyr-NLJ1B.v3.hap1.softmasked.fasta /home/miclark/MpyrAdapt/data/reference/Mpyr-NLJ1B.v3.hap1.softmasked.fasta

```

Run FastQC on raw reads, run MultiQC to summarize results, and check percent adapter dimer
```{bash, eval = FALSE}
sbatch ./scripts/read_QC.sbatch "/home/miclark/MpyrAdapt/data/raw" "raw_reads_QC" # Submitted batch job 109557, Submitted batch job 155663

```

Trim raw reads with Fastp
```{bash, eval = FALSE}
sbatch ./scripts/trim_reads.sbatch

# job finished fastp, then stopped running before running multiQC. Not sure if this is an issue with the node, with the script, or with the mutliQC module

# running multiQC manually: 
#run multiqc 
salloc --time=24:00:00 --mem=50Gb --ntasks=1 --cpus-per-task=1 --partition=lab-mpinsky \
        --qos=pi-mpinsky \
        --account=pi-mpinsky

OUTDIR=/home/miclark/MpyrAdapt/data/processed

module load multiqc

multiqc -v -p -ip -f --data-dir --data-format tsv --cl-config "max_table_rows: 3000" --filename trimmed_report --outdir $OUTDIR/QC $OUTDIR

# ran in < 2 minutes! 

```

## Align to reference genome

Make array key for alignment to reference genome
```{bash, eval = FALSE}
# code in make_alignment_key.sbatch
# script to make alignment array key to run with wrapper-align_to_genome.sh

KEY=/home/miclark/MpyrAdapt/scripts/alignment_key.txt
INDIR=/home/miclark/MpyrAdapt/data/processed/merged

ls $INDIR/*mrg.fq.gz | \
    sed -E 's/.*\///' | \
    sed -E 's/_mrg\.fq\.gz$//' | \
    while read line; do
        echo -e "$line\t$INDIR/${line}_r1.fq.gz\t$INDIR/${line}_r2.fq.gz\t$INDIR/${line}_mrg.fq.gz"
    done > $KEY

```

Install bwa on hummingbird––using bwa-mem2 for now, but bwa is installed in case I need to use aln in the future
```{bash, eval = FALSE}
module load miniconda3
conda create -n bwa
conda activate bwa 
conda install -c bioconda -c conda-forge bwa

```

Index reference genome with bwa
```{bash, eval = FALSE}
module load bwa-mem2/2.2.1
reference=/home/miclark/MpyrAdapt/data/reference/Mpyr-NLJ1B.v3.hap1.softmasked.fasta
bwa-mem2 index $reference

```

Start array job to align to genome
```{bash, eval = FALSE}
sbatch ./scripts/wrapper-align_to_genome.sh # submitted as array job 126960

# just do mem on merged and unmerged to begin with, if things look crazy, can compare mem and aln 

# I was taking up the entire reserach partition, so I cancelled arrays 68-188, will re-run just those arrays with a node limit. 
  # I also altered align_to_genome.sbatch so that it doesn't have to decompress read files multiple times to speed up jobs
  # submitted as array job 151608 and 151660

# running separate alignment job with CCGP_alignment_key.txt for CCGP data, submitted as array job 181599
```

Install qualimap
```{bash, eval = FALSE}
module load miniconda3

conda create -n qualimap -c bioconda -c conda-forge qualimap
conda activate qualimap

qualimap --help
```

Run qualimap on aligned BAM files to check mapping quality and coverage
```{bash, eval = FALSE}
./scripts/wrapper-run_qualimap.sh # jobid 152761
# running on CCGP data separately, job id 181803

# summarize qualimap output with multiQC 
sbatch ./scripts/summarize_qualimap.sbatch # jobid 154531, 182483 with CCGP data 
```

Use samtools depth to generate per-site coverage histogram to inform depth cut-offs
```{bash, eval = FALSE}
# Strategy: job will be an array job, where each task is a different chromosome, so the list of chromosomes will act as a array key. Each task will use all bams

# index .bam files (only need to happen once) from an interactive job
salloc --time=24:00:00 --mem=100Gb --ntasks=1 --cpus-per-task=10 --partition=lab-mpinsky \
        --qos=pi-mpinsky \
        --account=pi-mpinsky
module load samtools
indir="/home/miclark/MpyrAdapt/data/alignments"
samtools index -M -@ 10 ${indir}/*.bam 

# chrom_list.txt is a list of chromosome names to use as an array key for the samtools depth job, which will run one chromosome per task in an array job

./scripts/wrapper-bam_depth.sh 

# Read output files into R and plot coverage histograms to inform depth cut-offs for variant calling

sbatch ./scripts/make_depth_histograms.sbatch # jobid 194896

```

## Call SNPs
For this project, we will use a join genotype call and genotype liklihood approach. 
- Genotypes will be called with samples with >10X mean depth
- Genotype liklihoods will be calucated for all samples 
- Analyses with GLs can be confirmed with genotype calls 
- Approach: call SNPs with bcftools mpileup and bcftools call. Loci will be filtered in two flavors: one for liklihoods and one for calls 

```{bash, eval = FALSE}

./scripts/wrapper-call_snps.sh # jobid 194785

# [E::mpileup] the sequence "-s" not found: /home/miclark/MpyrAdapt/data/alignments/Mpyr_ALB_AM_11.rmdup.bam

```


